---
title: "Mice Protein Data - Ahmed et al."
date: "14/10/2019"
output: html_document
---

```{r setup, include=FALSE}
library(here)
library(tidyverse)
library(devtools)
library(scPCA)
library(elasticnet)
library(Rtsne)
library(umap)
library(ggpubr)
library(naniar)
library(SingleCellExperiment)
library(cluster)
library(xtable)

knitr::opts_chunk$set(echo = FALSE)
set.seed(734576345)
```


```{r mice_data_prep, message = FALSE}
# load the data
mice_df <- read_csv(here("analyses", "mice_data", "Data_Cortex_Nuclear.csv"))

# split into target and background
target_df <- mice_df %>%
  dplyr::filter(Treatment == "Saline",
                Behavior == "S/C",
                Genotype == "Control" || Genotype == "Ts65Dn") %>%
  select(-Genotype, -Behavior, -Treatment, -MouseID) %>%
  impute_median_if(is.numeric)
background_df <- mice_df %>%
  dplyr::filter(Treatment == "Saline",
                Behavior == "C/S",
                Genotype == "Control") %>%
  select(-Genotype, -Behavior, -Treatment, -MouseID, -class) %>%
  impute_median_all
```

Please see the results section of the manuscript for a description of the
experimental design.

### PCA

```{r mice_pca}
# perform PCA
pca_mice <- prcomp(target_df[, 1:77], center = TRUE, scale = TRUE)

# compute the average silhouette widths
group_mem <- if_else(as.data.frame(target_df[, 78]) == "c-SC-s", 1, 2)
pca_group_sil <- silhouette(group_mem, dist(pca_mice$x))

# plot the 2D rep using first 2 components
df <- data.frame(
  PC1 = pca_mice$x[, 1],
  PC2 = pca_mice$x[, 2],
  class = target_df$class
  ) %>%
  dplyr::mutate(
    class = if_else(
      class == "c-SC-s",
      paste0(
        "Con. (",
        sprintf("%.3f", round(summary(pca_group_sil)$clus.avg.widths[1], 3)),
        ")"),
      paste0(
        "Tri. (",
        sprintf("%.3f", round(summary(pca_group_sil)$clus.avg.widths[2], 3)),
        ")"),
    )
  )
      
p_pca <- ggplot(df, aes(x = PC1, y = PC2, colour = class)) +
  ggtitle("PCA") +
  geom_point(alpha = 0.7) +
  scale_colour_viridis_d(name = "Mice",
                         begin = 0.1, end = 0.9) +
  theme_minimal()
p_pca
```


### cPCA

```{r cPCA_mice}
# run cPCA for using 40 logarithmically seperated contrastive param values
cpca_mice <- scPCA(target = target_df[, 1:77],
                   background = background_df,
                   penalties = 0,
                   n_centers = 2,
                   scale = TRUE,
                   max_iter = 1000)

# compute the average silhouette widths
cpca_group_sil <- silhouette(group_mem, dist(cpca_mice$x))

# create a dataframe to be plotted
cpca_df <- cpca_mice$x %>%
  as.data.frame() %>%
  dplyr::mutate(class = target_df$class) %>%
  dplyr::mutate(
    class = if_else(
      class == "c-SC-s",
      paste0(
        "Con. (",
        sprintf("%.3f", round(summary(cpca_group_sil)$clus.avg.widths[1], 3)),
        ")"),
      paste0(
        "Tri. (",
        sprintf("%.3f", round(summary(cpca_group_sil)$clus.avg.widths[2], 3)),
        ")"),
    )
  )

colnames(cpca_df) <- c("cPC1", "cPC2", "label")

# plot the results
p_cpca <- ggplot(cpca_df, aes(x = -cPC1, y = -cPC2, colour = label)) +
  geom_point(alpha = 0.7) +
  scale_colour_viridis_d(name = "Mice",
                         begin = 0.1, end = 0.9) +
  ggtitle("cPCA") +
  xlab("cPC1") +
  ylab("cPC2") +
  theme_minimal()
p_cpca


# run the code for 2-5 centers
cpca_diff_centers <- lapply(
  2:5,
  function(x){
    scPCA(target = target_df[, 1:77], background = background_df, penalties = 0,
          n_centers = x, scale = TRUE, max_iter = 1000)
  })
```


```{r cpca_diff_centers}
# this is done one-by-one to ensure the plots' axes are comparable

# two centers:
cpca_2_group_sil <- silhouette(group_mem, dist(cpca_diff_centers[[1]]$x))
cpca_2_df <- cpca_diff_centers[[1]]$x %>%
  as.data.frame() %>%
  dplyr::mutate(class = target_df$class) %>%
  dplyr::mutate(
    class = if_else(
      class == "c-SC-s",
      paste0(
        "Con. (",
        sprintf("%.3f", round(summary(cpca_2_group_sil)$clus.avg.widths[1], 3)),
        ")"),
      paste0(
        "Tri. (",
        sprintf("%.3f", round(summary(cpca_2_group_sil)$clus.avg.widths[2], 3)),
        ")"),
    )
  )

colnames(cpca_2_df) <- c("cPC1", "cPC2", "label")

# plot the results
p_cpca_2 <- ggplot(cpca_2_df, aes(x = -cPC1, y = -cPC2, colour = label)) +
  geom_point(alpha = 0.7) +
  scale_colour_viridis_d(name = "Mice",
                         begin = 0.1, end = 0.9) +
  ggtitle("Two Clusters") +
  xlab("cPC1") +
  ylab("cPC2") +
  theme_minimal()

# three centers:
cpca_3_group_sil <- silhouette(group_mem, dist(cpca_diff_centers[[2]]$x))
cpca_3_df <- cpca_diff_centers[[2]]$x %>%
  as.data.frame() %>%
  dplyr::mutate(class = target_df$class) %>%
  dplyr::mutate(
    class = if_else(
      class == "c-SC-s",
      paste0(
        "Con. (",
        sprintf("%.3f", round(summary(cpca_3_group_sil)$clus.avg.widths[1], 3)),
        ")"),
      paste0(
        "Tri. (",
        sprintf("%.3f", round(summary(cpca_3_group_sil)$clus.avg.widths[2], 3)),
        ")"),
    )
  )

colnames(cpca_3_df) <- c("cPC1", "cPC2", "label")

# plot the results
p_cpca_3 <- ggplot(cpca_3_df, aes(x = cPC1, y = cPC2, colour = label)) +
  geom_point(alpha = 0.7) +
  scale_colour_viridis_d(name = "Mice",
                         begin = 0.1, end = 0.9) +
  ggtitle("Three Clusters") +
  xlab("cPC1") +
  ylab("cPC2") +
  theme_minimal()

# four centers:
cpca_4_group_sil <- silhouette(group_mem, dist(cpca_diff_centers[[3]]$x))
cpca_4_df <- cpca_diff_centers[[3]]$x %>%
  as.data.frame() %>%
  dplyr::mutate(class = target_df$class) %>%
  dplyr::mutate(
    class = if_else(
      class == "c-SC-s",
      paste0(
        "Con. (",
        sprintf("%.3f", round(summary(cpca_4_group_sil)$clus.avg.widths[1], 3)),
        ")"),
      paste0(
        "Tri. (",
        sprintf("%.3f", round(summary(cpca_4_group_sil)$clus.avg.widths[2], 3)),
        ")"),
    )
  )

colnames(cpca_4_df) <- c("cPC1", "cPC2", "label")

# plot the results
p_cpca_4 <- ggplot(cpca_4_df, aes(x = cPC1, y = cPC2, colour = label)) +
  geom_point(alpha = 0.7) +
  scale_colour_viridis_d(name = "Mice",
                         begin = 0.1, end = 0.9) +
  ggtitle("Four Clusters") +
  xlab("cPC1") +
  ylab("cPC2") +
  theme_minimal()

# five centers:
cpca_5_group_sil <- silhouette(group_mem, dist(cpca_diff_centers[[4]]$x))
cpca_5_df <- cpca_diff_centers[[4]]$x %>%
as.data.frame() %>%
  dplyr::mutate(class = target_df$class) %>%
  dplyr::mutate(
    class = if_else(
      class == "c-SC-s",
      paste0(
        "Con. (",
        sprintf("%.3f", round(summary(cpca_5_group_sil)$clus.avg.widths[1], 3)),
        ")"),
      paste0(
        "Tri. (",
        sprintf("%.3f", round(summary(cpca_5_group_sil)$clus.avg.widths[2], 3)),
        ")"),
    )
  )

colnames(cpca_5_df) <- c("cPC1", "cPC2", "label")

# plot the results
p_cpca_5 <- ggplot(cpca_5_df, aes(x = cPC1, y = -cPC2, colour = label)) +
  geom_point(alpha = 0.7) +
  scale_colour_viridis_d(name = "Mice",
                         begin = 0.1, end = 0.9) +
  ggtitle("Five Clusters") +
  xlab("cPC1") +
  ylab("cPC2") +
  theme_minimal()

# combine the 4 plots
ggarrange(p_cpca_2, p_cpca_3, p_cpca_4, p_cpca_5,
          ncol = 2, nrow = 2)
```


### scPCA

```{r scPCA_mice, cache=TRUE}
# run scPCA for using 40 logarithmically seperated contrastive parameter values
# and possible 20 L1 penalty terms
scpca_mice <- scPCA(target = target_df[, 1:77],
                    background = background_df,
                    n_centers = 2,
                    scale = TRUE,
                    max_iter = 1000)


# compute the average silhouette widths
scpca_group_sil <- silhouette(group_mem, dist(scpca_mice$x))

# create a dataframe to be plotted
scpca_df <- scpca_mice$x %>%
  as.data.frame() %>%
  dplyr::mutate(class = target_df$class) %>%
  dplyr::mutate(
    class = if_else(
      class == "c-SC-s",
      paste0(
        "Con. (",
        sprintf("%.3f", round(summary(scpca_group_sil)$clus.avg.widths[1], 3)),
        ")"),
      paste0(
        "Tri. (",
        sprintf("%.3f", round(summary(scpca_group_sil)$clus.avg.widths[2], 3)),
        ")"),
    )
  )

colnames(scpca_df) <- c("scPC1", "scPC2", "label")

# plot the results
p_scpca <- ggplot(scpca_df, aes(x = -scPC1, y = scPC2, colour = label)) +
  geom_point(alpha = 0.7) +
  scale_colour_viridis_d(name = "Mice",
                         begin = 0.1, end = 0.9) +
  ggtitle("scPCA") +
  xlab("scPC1") +
  ylab("scPC2") +
  theme_minimal()
p_scpca

# create table of proteins with non-zero weights in first and second loadings
prot_scpca1 <- data.frame(
  "name" = str_replace(names(scpca_mice$rotation[, 1]), "_N", ""),
  "weight" = scpca_mice$rotation[, 1]
) %>%
  dplyr::filter(weight != 0) %>%
  xtable(
    caption = "Proteins with non-zero weights in first scPCA loadings vector.",
    digits = 4
  )
prot_scpca2 <- data.frame(
  "name" = str_replace(names(scpca_mice$rotation[, 2]), "_N", ""),
  "weight" = scpca_mice$rotation[, 2]
) %>%
  dplyr::filter(weight != 0) %>%
  xtable(
    caption = "Proteins with non-zero weights in second scPCA loadings vector.",
    digits = 4
  )
# run the code for 2-5 centers
scpca_diff_centers <- lapply(
  2:5,
  function(x){
    scPCA(target = target_df[, 1:77], background = background_df,
          n_centers = x, scale = TRUE, max_iter = 1000)
  })
```

```{r scpca_diff_centers}
# this is done one-by-one to ensure the plots' axes are comparable

# two centers:
scpca_2_group_sil <- silhouette(group_mem, dist(scpca_diff_centers[[1]]$x))
scpca_2_df <- scpca_diff_centers[[1]]$x %>%
  as.data.frame() %>%
  dplyr::mutate(class = target_df$class) %>%
  dplyr::mutate(
    class = if_else(
      class == "c-SC-s",
      paste0(
        "Con. (",
        sprintf("%.3f", round(summary(scpca_2_group_sil)$clus.avg.widths[1], 3)),
        ")"),
      paste0(
        "Tri. (",
        sprintf("%.3f", round(summary(scpca_2_group_sil)$clus.avg.widths[2], 3)),
        ")"),
    )
  )

colnames(scpca_2_df) <- c("scPC1", "scPC2", "label")

p_scpca_2 <- ggplot(scpca_2_df, aes(x = -scPC1, y = scPC2, colour = label)) +
  geom_point(alpha = 0.7) +
  scale_colour_viridis_d(name = "Mice",
                         begin = 0.1, end = 0.9) +
  ggtitle("Two Clusters") +
  xlab("scPC1") +
  ylab("scPC2") +
  theme_minimal()

# three centers:
scpca_3_group_sil <- silhouette(group_mem, dist(scpca_diff_centers[[2]]$x))
scpca_3_df <- scpca_diff_centers[[2]]$x %>%
  as.data.frame() %>%
  dplyr::mutate(class = target_df$class) %>%
  dplyr::mutate(
    class = if_else(
      class == "c-SC-s",
      paste0(
        "Con. (",
        sprintf("%.3f", round(summary(scpca_3_group_sil)$clus.avg.widths[1], 3)),
        ")"),
      paste0(
        "Tri. (",
        sprintf("%.3f", round(summary(scpca_3_group_sil)$clus.avg.widths[2], 3)),
        ")"),
    )
  )

colnames(scpca_3_df) <- c("scPC1", "scPC2", "label")

p_scpca_3 <- ggplot(scpca_3_df, aes(x = -scPC1, y = -scPC2, colour = label)) +
  geom_point(alpha = 0.7) +
  scale_colour_viridis_d(name = "Mice",
                         begin = 0.1, end = 0.9) +
  ggtitle("Three Clusters") +
  xlab("scPC1") +
  ylab("scPC2") +
  theme_minimal()

# four centers:
scpca_4_group_sil <- silhouette(group_mem, dist(scpca_diff_centers[[3]]$x))
scpca_4_df <- scpca_diff_centers[[3]]$x %>%
  as.data.frame() %>%
  dplyr::mutate(class = target_df$class) %>%
  dplyr::mutate(
    class = if_else(
      class == "c-SC-s",
      paste0(
        "Con. (",
        sprintf("%.3f", round(summary(scpca_4_group_sil)$clus.avg.widths[1], 3)),
        ")"),
      paste0(
        "Tri. (",
        sprintf("%.3f", round(summary(scpca_4_group_sil)$clus.avg.widths[2], 3)),
        ")"),
    )
  )

colnames(scpca_4_df) <- c("scPC1", "scPC2", "label")

p_scpca_4 <- ggplot(scpca_4_df, aes(x = -scPC1, y = -scPC2, colour = label)) +
  geom_point(alpha = 0.7) +
  scale_colour_viridis_d(name = "Mice",
                         begin = 0.1, end = 0.9) +
  ggtitle("Four Clusters") +
  xlab("scPC1") +
  ylab("scPC2") +
  theme_minimal()

# five centers:
scpca_5_group_sil <- silhouette(group_mem, dist(scpca_diff_centers[[4]]$x))
scpca_5_df <- scpca_diff_centers[[4]]$x %>%
  as.data.frame() %>%
  dplyr::mutate(class = target_df$class) %>%
  dplyr::mutate(
    class = if_else(
      class == "c-SC-s",
      paste0(
        "Con. (",
        sprintf("%.3f", round(summary(scpca_5_group_sil)$clus.avg.widths[1], 3)),
        ")"),
      paste0(
        "Tri. (",
        sprintf("%.3f", round(summary(scpca_5_group_sil)$clus.avg.widths[2], 3)),
        ")"),
    )
  )

colnames(scpca_5_df) <- c("scPC1", "scPC2", "label")

p_scpca_5 <- ggplot(scpca_5_df, aes(x = -scPC1, y = -scPC2, colour = label)) +
  geom_point(alpha = 0.7) +
  scale_colour_viridis_d(name = "Mice",
                         begin = 0.1, end = 0.9) +
  ggtitle("Five Clusters") +
  xlab("scPC1") +
  ylab("scPC2") +
  theme_minimal()

# combine the 4 plots
ggarrange(p_scpca_2, p_scpca_3, p_scpca_4, p_scpca_5,
          ncol = 2, nrow = 2)
```


### Comparison of cPCA and scPCA Loadings

```{r loadings_comp, warning=FALSE, message=FALSE}
# compute the relative absolute loading ratios
cpca_loads_df <- cpca_mice$rotation %>%
  as.data.frame() %>%
  dplyr::mutate(
    V1 = abs(V1),
    V2 = abs(V2),
    V1 = (V1-min(V1))/(max(V1)-min(V1)),
    V2 = (V2-min(V2))/(max(V2)-min(V2))
  )
scpca_loads_df <- scpca_mice$rotation %>%
  as.data.frame() %>%
  dplyr::mutate(
    V1 = abs(V1),
    V2 = abs(V2),
    V1 = (V1-min(V1))/(max(V1)-min(V1)),
    V2 = (V2-min(V2))/(max(V2)-min(V2))
  )

# create the loadings comparison plot
load_diff_df <- bind_rows(
  cpca_loads_df,
  scpca_loads_df
) %>%
  dplyr::mutate(
    sparse = c(rep("0", ncol(target_df[, 1:77])),
               rep("1", ncol(target_df[, 1:77]))),
    sparse = factor(sparse, labels = c("cPCA", "scPCA")),
    prot = rep(1:ncol(target_df[, 1:77]), 2)
  )

colnames(load_diff_df) <- c("comp1", "comp2", "sparse", "prot")

# order the genes based on decreasing rel abs cpca loadings
ord_1_prot <- factor(load_diff_df$prot,
                     levels = sort(load_diff_df$comp1[1:ncol(target_df[, 1:77])],
                                   decreasing = TRUE, index.return = TRUE)$ix)
ord_2_prot <- factor(load_diff_df$prot,
                     levels = sort(load_diff_df$comp2[1:ncol(target_df[, 1:77])],
                                   decreasing = TRUE, index.return = TRUE)$ix)
load_diff_df <- load_diff_df %>%
  dplyr::mutate(
    prot1 = ord_1_prot,
    prot2 = ord_2_prot
  )

p1 <- load_diff_df %>% 
  ggplot(aes(y = comp1, x = prot1,
             fill = sparse, alpha = sparse)) +
    geom_bar(stat = "identity", position = position_identity(), width = 1) +
    xlab("Proteins") +
    ylab("Relative Abs. Weight") +
    ggtitle("First Loadings Vector Comparison") +
    scale_fill_viridis_d(name = "Method",
                         labels = c("cPCA", "scPCA"),
                         begin = 0.9, end = 0.1) +
    scale_alpha_discrete(range = c(0.5, 1),
                         guide = FALSE) +
    theme_minimal() +
    theme(panel.grid.major = element_blank(),
          panel.grid.minor = element_blank(),
          axis.text.x=element_blank(),
          axis.ticks.x=element_blank())

p2 <- load_diff_df %>%
  ggplot(aes(y = comp2, x = prot2,
             fill = sparse, alpha = sparse)) +
    geom_bar(stat = "identity", position = position_identity(), width = 1) +
    xlab("Proteins") +
    ylab("Relative Abs. Weight") +
    ggtitle("Second Loadings Vector Comparison") +
    scale_fill_viridis_d(name = "Method",
                         labels = c("cPCA", "scPCA"),
                         begin = 0.9, end = 0.1) +
    scale_alpha_discrete(range = c(0.5, 1),
                         guide = FALSE) +
    theme_minimal() +
    theme(panel.grid.major = element_blank(),
          panel.grid.minor = element_blank(),
          axis.text.x=element_blank(),
          axis.ticks.x=element_blank())

loadings_comp <- annotate_figure(
  ggarrange(p1, p2, nrow = 1, ncol = 2,
            common.legend = TRUE, legend = "right")
)
loadings_comp
```


### cPCA (Tuned with Cross-Validation)

5-fold cross-validation was used to tune the contrastive parameter.

```{r cPCA_cv_mice}
# run cPCA for using 40 logarithmically seperated contrastive param values
cpca_cv_mice <- scPCA(target = target_df[, 1:77],
                      background = background_df,
                      penalties = 0,
                      n_centers = 2,
                      scale = TRUE,
                      cv = 5,
                      max_iter = 1000)

# compute the average silhouette widths
cpca_cv_group_sil <- silhouette(group_mem, dist(cpca_cv_mice$x))

# create a dataframe to be plotted
cpca_cv_df <- cpca_cv_mice$x %>%
  as.data.frame() %>%
  dplyr::mutate(class = target_df$class) %>%
  dplyr::mutate(
    class = if_else(
      class == "c-SC-s",
      paste0(
        "Con. (",
        sprintf("%.3f", round(summary(cpca_cv_group_sil)$clus.avg.widths[1], 3)),
        ")"),
      paste0(
        "Tri. (",
        sprintf("%.3f", round(summary(cpca_cv_group_sil)$clus.avg.widths[2], 3)),
        ")"),
    )
  )

colnames(cpca_cv_df) <- c("cPC1", "cPC2", "label")

# plot the results
p_cpca_cv <- ggplot(cpca_cv_df, aes(x = -cPC1, y = -cPC2, colour = label)) +
  geom_point(alpha = 0.7) +
  scale_colour_viridis_d(name = "Mice",
                         begin = 0.1, end = 0.9) +
  ggtitle("5-Fold CV cPCA") +
  xlab("cPC1") +
  ylab("cPC2") +
  theme_minimal()
p_cpca_cv
```


### scPCA (Tuned with Cross-Validation)

5-fold cross-validation was used to tune the hyperparameters.

```{r scPCA_cv_mice, cache=TRUE}
# run cv scPCA for using 40 logarithmically seperated contrastive parameter
# values and possible 20 L1 penalty terms
scpca_cv_mice <- scPCA(target = target_df[, 1:77],
                    background = background_df,
                    n_centers = 2,
                    scale = TRUE,
                    cv = 5,
                    max_iter = 1000)


# compute the average silhouette widths
scpca_cv_group_sil <- silhouette(group_mem, dist(scpca_cv_mice$x))

# create a dataframe to be plotted
scpca_cv_df <- scpca_cv_mice$x %>%
  as.data.frame() %>%
  dplyr::mutate(class = target_df$class) %>%
  dplyr::mutate(
    class = if_else(
      class == "c-SC-s",
      paste0(
        "Con. (",
        sprintf("%.3f", round(summary(scpca_cv_group_sil)$clus.avg.widths[1], 3)),
        ")"),
      paste0(
        "Tri. (",
        sprintf("%.3f", round(summary(scpca_cv_group_sil)$clus.avg.widths[2], 3)),
        ")"),
    )
  )

colnames(scpca_cv_df) <- c("scPC1", "scPC2", "label")

# plot the results
p_scpca_cv <- ggplot(scpca_cv_df, aes(x = -scPC1, y = scPC2, colour = label)) +
  geom_point(alpha = 0.7) +
  scale_colour_viridis_d(name = "Mice",
                         begin = 0.1, end = 0.9) +
  ggtitle("5-Fold CV scPCA") +
  xlab("scPC1") +
  ylab("scPC2") +
  theme_minimal()
p_scpca_cv
```

### T-SNE

```{r t-sne_mice}
# run the tsne algorithm with default 
tsne_mice <- Rtsne(target_df[, 1:77], theta = 0, pca_scale = TRUE)

# compute the average silhouette widths
tsne_group_sil <- silhouette(group_mem, dist(tsne_mice$Y))

# create the t-sne dataframe
tsne_df <- tsne_mice$Y %>%
  as.data.frame() %>%
  dplyr::mutate(class = target_df$class) %>%
  dplyr::mutate(
    class = if_else(
      class == "c-SC-s",
      paste0(
        "Con. (",
        sprintf("%.3f", round(summary(tsne_group_sil)$clus.avg.widths[1], 3)),
        ")"),
      paste0(
        "Tri. (",
        sprintf("%.3f", round(summary(tsne_group_sil)$clus.avg.widths[2], 3)),
        ")"),
    )
  )
colnames(tsne_df) <- c("tSNE1", "tSNE2", "label")

# plot the result
p_tsne <- ggplot(tsne_df, aes(x = tSNE1, y = tSNE2, colour = label)) +
  geom_point(alpha = 0.7) +
  scale_colour_viridis_d(name = "Mice",
                         begin = 0.1, end = 0.9) +
  ggtitle("t-SNE") +
  xlab("tSNE1") +
  ylab("tSNE2") +
  theme_minimal()
p_tsne
```

### UMAP

```{r umap_mice}
# run the tsne algorithm with default 
umap_mice <- umap(target_df[, 1:77])


# compute the average silhouette widths
umap_group_sil <- silhouette(group_mem, dist(umap_mice$layout))

# create the t-sne dataframe
umap_df <- umap_mice$layout %>%
  as.data.frame() %>%
  dplyr::mutate(class = target_df$class) %>%
  dplyr::mutate(
    class = if_else(
      class == "c-SC-s",
      paste0(
        "Con. (",
        sprintf("%.3f", round(summary(umap_group_sil)$clus.avg.widths[1], 3)),
        ")"),
      paste0(
        "Tri. (",
        sprintf("%.3f", round(summary(umap_group_sil)$clus.avg.widths[2], 3)),
        ")"),
    )
  )

colnames(umap_df) <- c("UMAP1", "UMAP2", "label")

# plot the results
p_umap <- ggplot(umap_df, aes(x = UMAP1, y = UMAP2, colour = label)) +
  geom_point(alpha = 0.7) +
  scale_colour_viridis_d(name = "Mice",
                         begin = 0.1, end = 0.9) +
  ggtitle("UMAP") +
  theme_minimal()
p_umap
```

### Combined Plot

```{r result_mice, fig.retina=3, fig.height=8, fig.width=8}
ggarrange(
  ggarrange(p_pca, p_umap, p_cpca, p_scpca,
            ncol = 2, nrow = 2),
  loadings_comp,
  nrow = 2,
  labels = "AUTO",
  heights = c(2, 1)
)
```