---
title: "Bone Marrow Mononuclear Cells - Zheng et al."
author: "Philippe Boileau"
date: "15/10/2019"
output: html_document
---

```{r setup, include=FALSE}
library(here)
library(tidyverse)
library(scPCA)
library(Rtsne)
library(umap)
library(ggpubr)
library(SingleCellExperiment)
library(cluster)

knitr::opts_chunk$set(echo = FALSE)
```

# Data

scRNA-seq data was captured from the bone marrow mononuclear cells of two
patients with leukemia before and after undergoing a hematopoietic stem cell
transplant. We subjected the cells from each patient to various dimensionality
reduction techniques to determine whether they are capable of distinguishing the
treatment states of the cells. scRNA-seq data from two healthy controls are used
to perform cPCA and scPCA.

```{r load_data, message=FALSE, warning=FALSE}
# load the data
source(file = here("analyses/bmmc_data/helpers/load_count_data.R"))

# retain genes that contain 5 or more non-zero counts accross all cells
sce <- sce[(Matrix::rowSums(counts(sce) != 0) > 4), ]

# retain 1000 most variable genes
vars <- rowVars(as.matrix(log1p(counts(sce))))
names(vars) <- rownames(sce)
vars <- sort(vars, decreasing = TRUE)
core <- sce[names(vars)[1:1000], ]

# split into target and background datasets
background_sce <- core[, which(core$cell_prov %in% c("healthy_1", "healthy_2"))]
target_sce <- core[, which(!(core$cell_prov %in% c("healthy_1", "healthy_2")))]
patient_027_sce <- core[, which(core$cell_prov %in%
                                  c("pre_trans_027", "post_trans_027"))]
patient_035_sce <- core[, which(core$cell_prov %in%
                                  c("pre_trans_035", "post_trans_035"))]

background <- t(counts(background_sce))
target <- t(counts(target_sce))
patient_027 <- t(counts(patient_027_sce))
patient_035 <- t(counts(patient_035_sce))

# get the classes
bmmc_class <- target_sce$cell_prov %>% factor
pat027_class <- patient_027_sce$cell_prov %>% factor
pat035_class <- patient_035_sce$cell_prov %>% factor
```


# AML027

## PCA

```{r pca_pat027}
# perform PCA
bmmc_pca <- prcomp(patient_027, center = TRUE, scale. = TRUE)

# compute the average silhouette widths
group_mem <- if_else(pat027_class == "pre_trans_027", 1, 2)
pca_group_sil <- silhouette(group_mem, dist(bmmc_pca$x))

# plot the 2D representation
pca_df <- data.frame(
  PC1 = bmmc_pca$x[, 1],
  PC2 = bmmc_pca$x[, 2],
  class = factor(pat027_class, levels = c("pre_trans_027", "post_trans_027"))
) %>%
  dplyr::mutate(
    class = if_else(
      class == "pre_trans_027",
      paste0(
        "Pre (",
        sprintf("%.3f", round(summary(pca_group_sil)$clus.avg.widths[1], 3)),
        ")"),
      paste0(
        "Post (",
        sprintf("%.3f", round(summary(pca_group_sil)$clus.avg.widths[2], 3)),
        ")"),
    )
  )
pca_p <- pca_df %>%
  ggplot(aes(x = PC1, y = PC2, colour = class)) +
  geom_point(alpha = 0.5) +
  ggtitle("PCA") +
  scale_colour_viridis_d(name = "AML027",
                         begin = 0.1, end = 0.9) +
  theme_minimal()
pca_p
```


## cPCA

```{r cpca_pat027}
set.seed(1436262)

# perform cpca
bmmc_cpca <- scPCA(as.matrix(patient_027), as.matrix(background), center = TRUE,
                   scale = TRUE, penalties = 0, n_centers = 2, max_iter = 1000)

# compute the average silhouette widths
cpca_group_sil <- silhouette(group_mem, dist(bmmc_cpca$x))

# plot the 2D representation
cpca_df <- data.frame(
  cPC1 = bmmc_cpca$x[, 1],
  cPC2 = bmmc_cpca$x[, 2],
  class = factor(pat027_class, levels = c("pre_trans_027", "post_trans_027"))
) %>%
  dplyr::mutate(
    class = if_else(
      class == "pre_trans_027",
      paste0(
        "Pre (",
        sprintf("%.3f", round(summary(cpca_group_sil)$clus.avg.widths[1], 3)),
        ")"),
      paste0(
        "Post (",
        sprintf("%.3f", round(summary(cpca_group_sil)$clus.avg.widths[2], 3)),
        ")"),
    )
  )

cpca_p <- cpca_df %>%
  ggplot(aes(x = cPC1, y = cPC2, colour = class)) +
  geom_point(alpha = 0.5) +
  ggtitle("cPCA") +
  scale_colour_viridis_d(name = "AML027",
                         begin = 0.1, end = 0.9) +
  theme_minimal()
cpca_p
```


## scPCA

```{r scpca_pat027}
# load results from cluster files
bmmc_scpca <- readRDS(here("analyses/bmmc_data/data/bmmc_scpca_027.rds"))

# compute the average silhouette widths
scpca_group_sil <- silhouette(group_mem, dist(bmmc_scpca$x))

# plot the 2D representation
scpca_df <- data.frame(
  scPC1 = bmmc_scpca$x[, 1],
  scPC2 = bmmc_scpca$x[, 2],
  class = factor(pat027_class, levels = c("pre_trans_027", "post_trans_027"))
) %>%
  dplyr::mutate(
    class = if_else(
      class == "pre_trans_027",
      paste0(
        "Pre (",
        sprintf("%.3f", round(summary(scpca_group_sil)$clus.avg.widths[1], 3)),
        ")"),
      paste0(
        "Post (",
        sprintf("%.3f", round(summary(scpca_group_sil)$clus.avg.widths[2], 3)),
        ")"),
    )
  )

scpca_p <- scpca_df %>%
  ggplot(aes(x = scPC1, y = scPC2, colour = class)) +
  geom_point(alpha = 0.5) +
  ggtitle("scPCA") +
  scale_colour_viridis_d(name = "AML027",
                         begin = 0.1, end = 0.9) +
  theme_minimal()
scpca_p
```

### Analysis of Genes with Non-Zero Entries in Loadings Matrix

The following table provides the gene symbol and loadings of the genes with
non-zero entries in one of the loadings vectors of the first two scPCs.

```{r non_zer_bmmc_pat027}
rot_pat027_df <- bmmc_scpca$rotation %>%
  as_tibble %>%
  dplyr::mutate(
    gene_sym = rowData(patient_027_sce)$X2,
    scPC1 = V1,
    scPC2 = V2
  ) %>%
  select(-V1, -V2) %>%
  dplyr::filter(scPC1 != 0 | scPC2 != 0)
rot_pat027_df
```

### Comparison of Loadings Weights: cPCA and scPCA

The absolute valules of the loadings are not compared for this patient, since
only three entries in the leading loadings vectors produced by scPCA are
non-zero.


## cPCA (Tuned with Cross-Validation)

5-fold cross-validation was used to tune the contrastive parameter.

```{r cpca_cv_aml027}
# perform cv cpca
bmmc_cpca_cv <- readRDS(file = here("analyses/bmmc_data/data/bmmc_cpca_cv_027.rds"))

# compute the average silhouette widths
cpca_cv_group_sil <- silhouette(group_mem, dist(bmmc_cpca_cv$x))

# plot the 2D representation
cpca_cv_df <- data.frame(
  cPC1 = bmmc_cpca_cv$x[, 1],
  cPC2 = bmmc_cpca_cv$x[, 2],
  class = factor(pat027_class, levels = c("pre_trans_027", "post_trans_027"))
) %>%
  dplyr::mutate(
    class = if_else(
      class == "pre_trans_027",
      paste0(
        "Pre (",
        sprintf("%.3f", round(summary(cpca_cv_group_sil)$clus.avg.widths[1], 3)),
        ")"),
      paste0(
        "Post (",
        sprintf("%.3f", round(summary(cpca_cv_group_sil)$clus.avg.widths[2], 3)),
        ")"),
    )
  )

cpca_cv_p <- cpca_cv_df %>%
  ggplot(aes(x = cPC1, y = cPC2, colour = class)) +
  geom_point(alpha = 0.5) +
  ggtitle("cPCA") +
  scale_colour_viridis_d(name = "AML027",
                         begin = 0.1, end = 0.9) +
  theme_minimal()
cpca_cv_p
```



## scPCA (Tuned with Cross-Validation)

5-fold cross-validation was used to tune the contrastive parameter.

```{r scpca_cv_aml027}
# perform cv cpca
bmmc_scpca_cv <- readRDS(file = here("analyses/bmmc_data/data/bmmc_scpca_cv_027.rds"))

# compute the average silhouette widths
scpca_cv_group_sil <- silhouette(group_mem, dist(bmmc_scpca_cv$x))

# plot the 2D representation
scpca_cv_df <- data.frame(
  scPC1 = bmmc_scpca_cv$x[, 1],
  scPC2 = bmmc_scpca_cv$x[, 2],
  class = factor(pat027_class, levels = c("pre_trans_027", "post_trans_027"))
) %>%
  dplyr::mutate(
    class = if_else(
      class == "pre_trans_027",
      paste0(
        "Pre (",
        sprintf("%.3f", round(summary(scpca_cv_group_sil)$clus.avg.widths[1], 3)),
        ")"),
      paste0(
        "Post (",
        sprintf("%.3f", round(summary(scpca_cv_group_sil)$clus.avg.widths[2], 3)),
        ")"),
    )
  )

scpca_cv_p <- scpca_cv_df %>%
  ggplot(aes(x = scPC1, y = scPC2, colour = class)) +
  geom_point(alpha = 0.5) +
  ggtitle("scPCA") +
  scale_colour_viridis_d(name = "AML027",
                         begin = 0.1, end = 0.9) +
  theme_minimal()
scpca_cv_p
```


## t-SNE

### With Initial PCA Step

```{r tsne_pat027_pca, cache=TRUE}
# get tsne representation
bmmc_tsne_pca <- Rtsne(as.matrix(patient_027), perplexity = 30, max_iter = 1000,
                       pca = TRUE, theta = 0)

# compute the average silhouette widths
tsne_pca_group_sil <- silhouette(group_mem, dist(bmmc_tsne_pca$Y))

# plot the first two axes
tsne_pca_df <- data.frame(
  TSNE1 = bmmc_tsne_pca$Y[, 1],
  TSNE2 = bmmc_tsne_pca$Y[, 2],
  class = factor(pat027_class, levels = c("pre_trans_027", "post_trans_027"))
) %>%
  dplyr::mutate(
    class = if_else(
      class == "pre_trans_027",
      paste0(
        "Pre (",
        sprintf("%.3f", round(summary(tsne_pca_group_sil)$clus.avg.widths[1], 3)),
        ")"),
      paste0(
        "Post (",
        sprintf("%.3f", round(summary(tsne_pca_group_sil)$clus.avg.widths[2], 3)),
        ")"),
    )
  )

tsne_pca_p <- tsne_pca_df %>%
  ggplot(aes(x = TSNE1, y = TSNE2, colour = class)) +
  geom_point(alpha = 0.5) +
  ggtitle("t-SNE") +
  xlab("tSNE1") +
  ylab("tSNE2") +
  scale_colour_viridis_d(name = "AML027",
                         begin = 0.1, end = 0.9) +
  theme_minimal()
tsne_pca_p
```

### Without Initial PCA Step

```{r tsne_pat027, cache=TRUE}
# get tsne representation
bmmc_tsne <- Rtsne(as.matrix(patient_027), perplexity = 30, max_iter = 1000,
                   pca = FALSE, theta = 0)

# compute the average silhouette widths
tsne_group_sil <- silhouette(group_mem, dist(bmmc_tsne$Y))

# plot the first two axes
tsne_df <- data.frame(
  TSNE1 = bmmc_tsne$Y[, 1],
  TSNE2 = bmmc_tsne$Y[, 2],
  class = factor(pat027_class, levels = c("pre_trans_027", "post_trans_027"))
) %>%
  dplyr::mutate(
    class = if_else(
      class == "pre_trans_027",
      paste0(
        "Pre (",
        sprintf("%.3f", round(summary(tsne_group_sil)$clus.avg.widths[1], 3)),
        ")"),
      paste0(
        "Post (",
        sprintf("%.3f", round(summary(tsne_group_sil)$clus.avg.widths[2], 3)),
        ")"),
    )
  )

tsne_p <- tsne_df %>%
  ggplot(aes(x = TSNE1, y = TSNE2, colour = class)) +
  geom_point(alpha = 0.5) +
  ggtitle("t-SNE") +
  xlab("tSNE1") +
  ylab("tSNE2") +
  scale_colour_viridis_d(name = "AML027",
                         begin = 0.1, end = 0.9) +
  theme_minimal()
tsne_p
```

## UMAP

```{r umap_pat027}
# get the umap representation
bmmc_umap <- umap(as.matrix(patient_027), n_neighors = 30, min_dist = 0.02)

# compute the average silhouette widths
umap_group_sil <- silhouette(group_mem, dist(bmmc_umap$layout))

# plot the first two axes
umap_df <- bmmc_umap$layout %>%
  as.data.frame() %>%
  dplyr::mutate(
    class = factor(pat027_class, levels = c("pre_trans_027", "post_trans_027"))
  ) %>%
  dplyr::mutate(
    class = if_else(
      class == "pre_trans_027",
      paste0(
        "Pre (",
        sprintf("%.3f", round(summary(umap_group_sil)$clus.avg.widths[1], 3)),
        ")"),
      paste0(
        "Post (",
        sprintf("%.3f", round(summary(umap_group_sil)$clus.avg.widths[2], 3)),
        ")"),
    )
  )
colnames(umap_df) <- c("UMAP1", "UMAP2", "class")

umap_p <- umap_df %>%
  ggplot(aes(x = UMAP1, y = UMAP2, colour = class)) +
  geom_point(alpha = 0.5) +
  ggtitle("UMAP") +
  scale_colour_viridis_d(name = "AML027",
                         begin = 0.1, end = 0.9) +
  theme_minimal()
umap_p
```


## ZINB-WaVE

```{r zinbwave_pat027}
# load the zinbwave model
bmmc_zinbwave_sce <- readRDS(here("analyses/bmmc_data/data/zinbwave_027.rds"))

# compute the average silhouette widths
zinbwave_group_sil <- silhouette(group_mem,
                                 dist(reducedDim(bmmc_zinbwave_sce)[, 1:2]))

# create a dataframe for plotting
zinbwave_df <- data.frame(
  ZW1 = reducedDim(bmmc_zinbwave_sce)[, 1],
  ZW2 = reducedDim(bmmc_zinbwave_sce)[, 2],
  class = factor(pat027_class, levels = c("pre_trans_027", "post_trans_027"))
) %>%
  dplyr::mutate(
    class = if_else(
      class == "pre_trans_027",
      paste0(
        "Pre (",
        sprintf("%.3f", round(summary(zinbwave_group_sil)$clus.avg.widths[1], 3)),
        ")"),
      paste0(
        "Post (",
        sprintf("%.3f", round(summary(zinbwave_group_sil)$clus.avg.widths[2], 3)),
        ")"),
    )
  )

zinbwave_p <- zinbwave_df %>%
  ggplot(aes(x = ZW1, y = ZW2, colour = class)) +
  geom_point(alpha = 0.5) +
  ggtitle("ZINB-WaVE") +
  scale_colour_viridis_d(name = "AML027",
                         begin = 0.1, end = 0.9) +
  theme_minimal()
zinbwave_p
```


## SIMLR

```{r simlr_pat_027}
# apply SIMLR algorithm to the target datasaet
bmmc_simlr <- readRDS(here("analyses/bmmc_data/data/bmmc_simlr_027.rds"))

# get the silhouette width of the batch and biological groups
simlr_group_sil <- silhouette(group_mem, dist(bmmc_simlr$ydata))

# plot the 2D representation
simlr_df <- data.frame(
  SIMLR1 = bmmc_simlr$ydata[, 1],
  SIMLR2 = bmmc_simlr$ydata[, 2],
  class = factor(pat027_class, levels = c("pre_trans_027", "post_trans_027"))
) %>%
  dplyr::mutate(
    class = if_else(
      class == "pre_trans_027",
      paste0("Pre (",
             sprintf("%.3f",
                     round(summary(simlr_group_sil)$clus.avg.widths[1], 3)),
             ")"),
      paste0("Post (",
             sprintf("%.3f",
                     round(summary(simlr_group_sil)$clus.avg.widths[2], 3)),
             ")"))
    )

simlr_p <- simlr_df %>%
  ggplot(aes(x = SIMLR1, y = SIMLR2, colour = class)) +
  geom_point(alpha = 0.5) +
  ggtitle("SIMLR") +
  scale_colour_viridis_d(name = "AML027",
                         begin = 0.1, end = 0.9) +
  theme_minimal()

ggsave(here("manuscript/figures/aml027_SIMLR.png"),
       width = 15, height = 10, units = "cm", dpi = 300)

simlr_p
```


## Combined Plots

```{r combined_pat027, fig.retina=3, fig.width=10, fig.height=10}
pat027_p <- ggarrange(pca_p, zinbwave_p, tsne_p, umap_p, cpca_p, scpca_p,
            ncol = 2, nrow = 3)
pat027_p
```


# AML035

## PCA

```{r pca_pat035}
# perform PCA
bmmc_pca <- prcomp(patient_035, center = TRUE, scale. = TRUE)

# compute the average silhouette widths
group_mem <- if_else(pat035_class == "pre_trans_035", 1, 2)
pca_group_sil <- silhouette(group_mem, dist(bmmc_pca$x))

# plot the 2D representation
pca_df <- data.frame(
  PC1 = bmmc_pca$x[, 1],
  PC2 = bmmc_pca$x[, 2],
  class = factor(pat035_class, levels = c("pre_trans_035", "post_trans_035"))
) %>%
  dplyr::mutate(
    class = if_else(
      class == "pre_trans_035",
      paste0(
        "Pre (",
        sprintf("%.3f", round(summary(pca_group_sil)$clus.avg.widths[1], 3)),
        ")"),
      paste0(
        "Post (",
        sprintf("%.3f", round(summary(pca_group_sil)$clus.avg.widths[2], 3)),
        ")"),
    )
  )

pca_p <- pca_df %>%
  ggplot(aes(x = PC1, y = PC2, colour = class)) +
  geom_point(alpha = 0.5) +
  ggtitle("PCA") +
  scale_colour_viridis_d(name = "AML035",
                         begin = 0.1, end = 0.9) +
  theme_minimal()
pca_p
```


## cPCA

```{r cpca_pat035}
set.seed(1436262)

# perform cpca
bmmc_cpca <- scPCA(as.matrix(patient_035), as.matrix(background),
                   center = TRUE, scale = TRUE, penalties = 0, n_centers = 2,
                   max_iter = 1000)

# compute the average silhouette widths
cpca_group_sil <- silhouette(group_mem, dist(bmmc_cpca$x))

# plot the 2D representation
cpca_df <- data.frame(
  cPC1 = bmmc_cpca$x[, 1],
  cPC2 = bmmc_cpca$x[, 2],
  class = factor(pat035_class, levels = c("pre_trans_035", "post_trans_035"))
) %>%
  dplyr::mutate(
    class = if_else(
      class == "pre_trans_035",
      paste0(
        "Pre (",
        sprintf("%.3f", round(summary(cpca_group_sil)$clus.avg.widths[1], 3)),
        ")"),
      paste0(
        "Post (",
        sprintf("%.3f", round(summary(cpca_group_sil)$clus.avg.widths[2], 3)),
        ")"),
    )
  )

cpca_p <- cpca_df %>%
  ggplot(aes(x = cPC1, y = cPC2, colour = class)) +
  geom_point(alpha = 0.5) +
  ggtitle("cPCA") +
  scale_colour_viridis_d(name = "AML035",
                         begin = 0.1, end = 0.9) +
  theme_minimal()
cpca_p
```


## scPCA


```{r scpca_pat035}
# load results from cluster files
bmmc_scpca <- readRDS(here("analyses/bmmc_data/data/bmmc_scpca_035.rds"))

# compute the average silhouette widths
scpca_group_sil <- silhouette(group_mem, dist(bmmc_scpca$x))

# plot the 2D representation
scpca_df <- data.frame(
  scPC1 = bmmc_scpca$x[, 1],
  scPC2 = bmmc_scpca$x[, 2],
  class = factor(pat035_class, levels = c("pre_trans_035", "post_trans_035"))
) %>%
  dplyr::mutate(
    class = if_else(
      class == "pre_trans_035",
      paste0(
        "Pre (",
        sprintf("%.3f", round(summary(scpca_group_sil)$clus.avg.widths[1], 3)),
        ")"),
      paste0(
        "Post (",
        sprintf("%.3f", round(summary(scpca_group_sil)$clus.avg.widths[2], 3)),
        ")"),
    )
  )

scpca_p <- scpca_df %>%
  ggplot(aes(x = scPC1, y = scPC2, colour = class)) +
  geom_point(alpha = 0.5) +
  ggtitle("scPCA") +
  scale_colour_viridis_d(name = "AML035",
                         begin = 0.1, end = 0.9) +
  theme_minimal()
scpca_p
```


### Analysis of Genes with Non-Zero Entries in Loadings Matrix

The following table provides the gene symbol and loadings of the genes with
non-zero entries in one of the loadings vectors of the first two scPCs.

```{r non_zer_bmmc_pat035}
rot_035_df <- bmmc_scpca$rotation %>%
  as_tibble %>%
  dplyr::mutate(
    gene_sym = rowData(patient_035_sce)$X2,
    scPC1 = V1,
    scPC2 = V2
  ) %>%
  select(-V1, -V2) %>%
  dplyr::filter(scPC1 != 0 | scPC2 != 0)
rot_035_df
```

### Comparison of Loadings Weights: cPCA and scPCA

```{r pat035_comp, message = FALSE, warning = FALSE}
# compute the relative absolute loading ratios
cpca_loads_df <- bmmc_cpca$rotation %>%
  as.data.frame() %>%
  dplyr::mutate(
    V1 = abs(V1),
    V2 = abs(V2),
    V1 = (V1-min(V1))/(max(V1)-min(V1)),
    V2 = (V2-min(V2))/(max(V2)-min(V2))
  )
scpca_loads_df <- bmmc_scpca$rotation %>%
  as.data.frame() %>%
  dplyr::mutate(
    V1 = abs(V1),
    V2 = abs(V2),
    V1 = (V1-min(V1))/(max(V1)-min(V1)),
    V2 = (V2-min(V2))/(max(V2)-min(V2))
  )

# create the loadings comparison plot
load_diff_df <- bind_rows(
  cpca_loads_df,
  scpca_loads_df
) %>%
  dplyr::mutate(
    sparse = c(rep("0", ncol(patient_035)),
               rep("1", ncol(patient_035))),
    sparse = factor(sparse, labels = c("cPCA", "scPCA")),
    gene = rep(1:ncol(patient_035), 2)
  )

colnames(load_diff_df) <- c("comp1", "comp2", "sparse", "gene")

# order the genes based on decreasing rel abs cpca loadings
ord_1_gene <- factor(load_diff_df$gene,
                     levels = sort(load_diff_df$comp1[1:ncol(patient_035)],
                                   decreasing = TRUE, index.return = TRUE)$ix)
ord_2_gene <- factor(load_diff_df$gene,
                     levels = sort(load_diff_df$comp2[1:ncol(patient_035)],
                                   decreasing = TRUE, index.return = TRUE)$ix)
load_diff_df <- load_diff_df %>%
  dplyr::mutate(
    gene1 = ord_1_gene,
    gene2 = ord_2_gene
  )

p1 <- load_diff_df %>%
  ggplot(aes(y = abs(comp1), x = gene1, fill = sparse, alpha = sparse)) +
  geom_bar(stat = "identity", position = position_identity(), width = 1) +
  xlab("Genes") +
  ylab("Relative Abs. Weight") +
  ggtitle("First Loadings Vector Comparison") +
  scale_fill_viridis_d(name = "Method",
                       labels = c("cPCA", "scPCA"),
                       begin = 0.9, end = 0.1) +
  scale_alpha_discrete(range = c(0.5, 1),
                       guide = FALSE) +
  theme_minimal() +
  theme(panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        axis.text.x=element_blank(),
        axis.ticks.x=element_blank())

p2 <- load_diff_df %>%
  ggplot(aes(y = abs(comp2), x = gene2, fill = sparse, alpha = sparse)) +
  geom_bar(stat = "identity", position = position_identity(), width = 1) +
  xlab("Genes") +
  ylab("Relative Abs. Weight") +
  ggtitle("Second Loadings Vector Comparison") +
  scale_fill_viridis_d(name = "Method",
                       labels = c("cPCA", "scPCA"),
                       begin = 0.9, end = 0.1) +
  scale_alpha_discrete(range = c(0.5, 1),
                       guide = FALSE) +
  theme_minimal() +
  theme(panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        axis.text.x=element_blank(),
        axis.ticks.x=element_blank())

loadings_comp <- annotate_figure(
  ggarrange(p1, p2, nrow = 1, ncol = 2,
            common.legend = TRUE, legend = "right")
)
loadings_comp
```


## cPCA (Tuned with Cross-Validation)

5-fold cross-validation was used to tune the contrastive parameter.

```{r cpca_cv_aml035}
# perform cv cpca
bmmc_cpca_cv <- readRDS(file = here("analyses/bmmc_data/data/bmmc_cpca_cv_035.rds"))

# compute the average silhouette widths
cpca_cv_group_sil <- silhouette(group_mem, dist(bmmc_cpca_cv$x))

# plot the 2D representation
cpca_cv_df <- data.frame(
  cPC1 = bmmc_cpca_cv$x[, 1],
  cPC2 = bmmc_cpca_cv$x[, 2],
  class = factor(pat035_class, levels = c("pre_trans_035", "post_trans_035"))
) %>%
  dplyr::mutate(
    class = if_else(
      class == "pre_trans_035",
      paste0(
        "Pre (",
        sprintf("%.3f", round(summary(cpca_cv_group_sil)$clus.avg.widths[1], 3)),
        ")"),
      paste0(
        "Post (",
        sprintf("%.3f", round(summary(cpca_cv_group_sil)$clus.avg.widths[2], 3)),
        ")"),
    )
  )

cpca_cv_p <- cpca_cv_df %>%
  ggplot(aes(x = cPC1, y = cPC2, colour = class)) +
  geom_point(alpha = 0.5) +
  ggtitle("cPCA") +
  scale_colour_viridis_d(name = "AML035",
                         begin = 0.1, end = 0.9) +
  theme_minimal()
cpca_cv_p
```


## scPCA (Tuned with Cross-Validation)

5-fold cross-validation was used to tune the contrastive parameter.

```{r scpca_cv_aml035}
# perform cv cpca
bmmc_scpca_cv <- readRDS(file = here("analyses/bmmc_data/data/bmmc_scpca_cv_035.rds"))

# compute the average silhouette widths
scpca_cv_group_sil <- silhouette(group_mem, dist(bmmc_scpca_cv$x))

# plot the 2D representation
scpca_cv_df <- data.frame(
  scPC1 = bmmc_scpca_cv$x[, 1],
  scPC2 = bmmc_scpca_cv$x[, 2],
  class = factor(pat035_class, levels = c("pre_trans_035", "post_trans_035"))
) %>%
  dplyr::mutate(
    class = if_else(
      class == "pre_trans_035",
      paste0(
        "Pre (",
        sprintf("%.3f", round(summary(scpca_cv_group_sil)$clus.avg.widths[1], 3)),
        ")"),
      paste0(
        "Post (",
        sprintf("%.3f", round(summary(scpca_cv_group_sil)$clus.avg.widths[2], 3)),
        ")"),
    )
  )

scpca_cv_p <- scpca_cv_df %>%
  ggplot(aes(x = scPC1, y = scPC2, colour = class)) +
  geom_point(alpha = 0.5) +
  ggtitle("scPCA") +
  scale_colour_viridis_d(name = "AML035",
                         begin = 0.1, end = 0.9) +
  theme_minimal()
scpca_cv_p
```

## t-SNE

### With Initial PCA Step

```{r tsne_pat035_pca, cache=TRUE}
# get tsne representation
bmmc_tsne_pca <- Rtsne(as.matrix(patient_035), perplexity = 30, max_iter = 1000,
                   pca = TRUE, theta = 0)

# compute the average silhouette widths
tsne_pca_group_sil <- silhouette(group_mem, dist(bmmc_tsne_pca$Y))

# plot the first two axes
tsne_pca_df <- data.frame(
  TSNE1 = bmmc_tsne_pca$Y[, 1],
  TSNE2 = bmmc_tsne_pca$Y[, 2],
  class = factor(pat035_class, levels = c("pre_trans_035", "post_trans_035"))
) %>%
  dplyr::mutate(
    class = if_else(
      class == "pre_trans_035",
      paste0(
        "Pre (",
        sprintf("%.3f", round(summary(tsne_pca_group_sil)$clus.avg.widths[1], 3)),
        ")"),
      paste0(
        "Post (",
        sprintf("%.3f", round(summary(tsne_pca_group_sil)$clus.avg.widths[2], 3)),
        ")"),
    )
  )


tsne_pca_p <- tsne_pca_df %>%
  ggplot(aes(x = TSNE1, y = TSNE2, colour = class)) +
  geom_point(alpha = 0.5) +
  ggtitle("t-SNE") +
  xlab("tSNE1") +
  ylab("tSNE2") +
  scale_colour_viridis_d(name = "AML035",
                         begin = 0.1, end = 0.9) +
  theme_minimal()
tsne_pca_p
```

### Without Initial PCA Step

```{r tsne_pat035, cache=TRUE}
# get tsne representation
bmmc_tsne <- Rtsne(as.matrix(patient_035), perplexity = 30, max_iter = 1000,
                   pca = FALSE, theta = 0)

# compute the average silhouette widths
tsne_group_sil <- silhouette(group_mem, dist(bmmc_tsne$Y))

# plot the first two axes
tsne_df <- data.frame(
  TSNE1 = bmmc_tsne$Y[, 1],
  TSNE2 = bmmc_tsne$Y[, 2],
  class = factor(pat035_class, levels = c("pre_trans_035", "post_trans_035"))
) %>%
  dplyr::mutate(
    class = if_else(
      class == "pre_trans_035",
      paste0(
        "Pre (",
        sprintf("%.3f", round(summary(tsne_group_sil)$clus.avg.widths[1], 3)),
        ")"),
      paste0(
        "Post (",
        sprintf("%.3f", round(summary(tsne_group_sil)$clus.avg.widths[2], 3)),
        ")"),
    )
  )


tsne_p <- tsne_df %>%
  ggplot(aes(x = TSNE1, y = TSNE2, colour = class)) +
  geom_point(alpha = 0.5) +
  ggtitle("t-SNE") +
  xlab("tSNE1") +
  ylab("tSNE2") +
  scale_colour_viridis_d(name = "AML035",
                         begin = 0.1, end = 0.9) +
  theme_minimal()
tsne_p
```

## UMAP

```{r umap_pat035}
# get the umap representation
bmmc_umap <- umap(as.matrix(patient_035), n_neighors = 30, min_dist = 0.02)

# compute the average silhouette widths
umap_group_sil <- silhouette(group_mem, dist(bmmc_umap$layout))

# plot the first two axes
umap_df <- bmmc_umap$layout %>%
  as.data.frame() %>%
  dplyr::mutate(
    class = factor(pat035_class, levels = c("pre_trans_035", "post_trans_035"))
  ) %>%
  dplyr::mutate(
    class = if_else(
      class == "pre_trans_035",
      paste0(
        "Pre (",
        sprintf("%.3f", round(summary(umap_group_sil)$clus.avg.widths[1], 3)),
        ")"),
      paste0(
        "Post (",
        sprintf("%.3f", round(summary(umap_group_sil)$clus.avg.widths[2], 3)),
        ")"),
    )
  )
colnames(umap_df) <- c("UMAP1", "UMAP2", "class")

umap_p <- umap_df %>%
  ggplot(aes(x = UMAP1, y = UMAP2, colour = class)) +
  geom_point(alpha = 0.5) +
  ggtitle("UMAP") +
  scale_colour_viridis_d(name = "AML035",
                         begin = 0.1, end = 0.9) +
  theme_minimal()
umap_p
```

## ZINB-WaVE

```{r zinbwave_pat035}
# load the zinbwave model
bmmc_zinbwave_sce <- readRDS(here("analyses/bmmc_data/data/zinbwave_035.rds"))

# compute the average silhouette widths
zinbwave_group_sil <- silhouette(group_mem,
                                 dist(reducedDim(bmmc_zinbwave_sce)[, 1:2]))

# create a dataframe for plotting
zinbwave_df <- data.frame(
  ZW1 = reducedDim(bmmc_zinbwave_sce)[, 1],
  ZW2 = reducedDim(bmmc_zinbwave_sce)[, 2],
  class = factor(pat035_class, levels = c("pre_trans_035", "post_trans_035"))
) %>%
  dplyr::mutate(
    class = if_else(
      class == "pre_trans_035",
      paste0(
        "Pre (",
        sprintf("%.3f", round(summary(zinbwave_group_sil)$clus.avg.widths[1], 3)),
        ")"),
      paste0(
        "Post (",
        sprintf("%.3f", round(summary(zinbwave_group_sil)$clus.avg.widths[2], 3)),
        ")"),
    )
  )

zinbwave_p <- zinbwave_df %>%
  ggplot(aes(x = ZW1, y = ZW2, colour = class)) +
  geom_point(alpha = 0.5) +
  ggtitle("ZINB-WaVE") +
  scale_colour_viridis_d(name = "AML035",
                         begin = 0.1, end = 0.9) +
  theme_minimal()
zinbwave_p
```

## SIMLR

```{r simlr_pat_035}
# apply SIMLR algorithm to the target datasaet
bmmc_simlr <- readRDS(here("analyses/bmmc_data/data/bmmc_simlr_035.rds"))

# get the silhouette width of the batch and biological groups
simlr_group_sil <- silhouette(group_mem, dist(bmmc_simlr$ydata))

# plot the 2D representation
simlr_df <- data.frame(
  SIMLR1 = bmmc_simlr$ydata[, 1],
  SIMLR2 = bmmc_simlr$ydata[, 2],
  class = factor(pat035_class, levels = c("pre_trans_035", "post_trans_035"))
) %>%
  dplyr::mutate(
    class = if_else(
      class == "pre_trans_035",
      paste0("Pre (",
             sprintf("%.3f",
                     round(summary(simlr_group_sil)$clus.avg.widths[1], 3)),
             ")"),
      paste0("Post (",
             sprintf("%.3f",
                     round(summary(simlr_group_sil)$clus.avg.widths[2], 3)),
             ")"))
    )

simlr_p <- simlr_df %>%
  ggplot(aes(x = SIMLR1, y = SIMLR2, colour = class)) +
  geom_point(alpha = 0.5) +
  ggtitle("SIMLR") +
  scale_colour_viridis_d(name = "AML035",
                         begin = 0.1, end = 0.9) +
  theme_minimal()

ggsave(here("manuscript/figures/aml035_SIMLR.png"),
       width = 15, height = 10, units = "cm", dpi = 300)

simlr_p
```

## Combined Plots

```{r combined_pat035, , fig.retina=3, fig.width=10, fig.height=12}
pat035_p <- ggarrange(pca_p, zinbwave_p, tsne_p, umap_p, cpca_p, scpca_p,
            ncol = 2, nrow = 3)
pat035_complete <- ggarrange(
  pat035_p,
  loadings_comp,
  nrow = 2,
  labels = "AUTO",
  heights = c(3, 1)
)
pat035_complete
```

```{r save_dengue_data_plot, eval=FALSE}
ggsave(
  filename = here("/manuscript/figures/dengue_results.png"),
  width = 178,
  height = 200,
  units = "mm",
  dpi = 350,
  plot = pat035_complete,
  scale = 1.5
)
```