---
title: "Systems biological analysis of immunity to dengue - Kwissa et al."
author: "Philippe Boileau"
date: "15/10/2019"
output: html_document
---

```{r setup, include=FALSE}
library(here)
library(tidyverse)
library(scPCA)
library(Rtsne)
library(umap)
library(ggpubr)
library(GEOquery)
library(genefilter)
library(cluster)
library(xtable)

knitr::opts_chunk$set(echo = FALSE)

set.seed(9765967)
```

# Prepare the Data

The microarray data is fetched from the GEO (accession number GSE51808). The
data consists of the transcriptome of blood samples from 56 Thai males, 28 of
which were hospitalized with severe dengue, 19 patients which were convalescent
4 weeks after contracting dengue and 9 healthy controls. Among the patients
with dengue, 18 had a fever and 10 had a hemorrhagic fever.

The data was filtered; only the 500 most variable genes were retained. The
transcriptome data of the 9 healthy controls were removed and used as a
background dataset for cPCA and scPCA.  The remaining samples consist of the
target data.

```{r load_data, message=FALSE, warning=FALSE}
# load the data, already log2 transformed
ges <- getGEO("GSE51808")$GSE51808_series_matrix.txt.gz

# load the gene names
# download table from https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GPL13158
gene_names <- read_tsv(here("analyses/dengue_data/GPL13158-5065.txt"),
                       skip = 16)

# keep the 500 most variable genes
var_filt_ges <- varFilter(ges, var.cutoff = 1-500/nrow(exprs(ges)))

# extract the target and background datasets
control_label <- which(var_filt_ges$`status:ch1` == "control")
target <- t(exprs(var_filt_ges)[, -control_label])
background <- t(exprs(var_filt_ges)[, control_label])

# get the target data labels
dengue_class <- var_filt_ges$`status:ch1`[-control_label]
```

# Dimension Reduction Methods

The target (and where appropriate the background) were subjected to various
dimension reduction methods. Ideally, three clusters should be found in the
target dataset: patients with dengue fever, with dengue hemorrhagic fever, and
patients convalescent after 4 weeks.


## PCA

```{r pca}
# perform PCA
dengue_pca <- prcomp(target, center = TRUE, scale. = FALSE)

# compute the average silhouette widths
group_mem <- if_else(dengue_class == "convalescent", 1,
                     if_else(dengue_class == "DF", 2, 3))
pca_group_sil <- silhouette(group_mem, dist(dengue_pca$x))

# plot the 2D representation
pca_df <- data.frame(
  PC1 = dengue_pca$x[, 1],
  PC2 = dengue_pca$x[, 2],
  class = dengue_class
) %>%
  dplyr::mutate(
    class = if_else(
      class == "convalescent",
      paste0(
        "Conv. (",
        sprintf("%.3f", round(summary(pca_group_sil)$clus.avg.widths[1], 3)),
        ")"),
      if_else(
        class == "DF",
        paste0(
        "DF (",
        sprintf("%.3f", round(summary(pca_group_sil)$clus.avg.widths[2], 3)),
        ")"),
        paste0("DHF (",
        sprintf("%.3f", round(summary(pca_group_sil)$clus.avg.widths[3], 3)),
        ")")
      )
    )
  )

pca_p <- pca_df %>%
  ggplot(aes(x = PC1, y = PC2, colour = class)) +
  geom_point() +
  ggtitle("PCA") +
  scale_colour_viridis_d(name = "Class", 
                         begin = 0.1, end = 0.9) +
  theme_minimal()
pca_p
```


## cPCA

```{r cpca}
# perform cpca
dengue_cpca <- scPCA(target, background, center = TRUE,
                     penalties = 0, n_centers = 3, max_iter = 1000)

# compute the average silhouette widths
cpca_group_sil <- silhouette(group_mem, dist(dengue_cpca$x))

# plot the 2D representation
cpca_df <- data.frame(
  cPC1 = -dengue_cpca$x[, 1],
  cPC2 = dengue_cpca$x[, 2],
  class = dengue_class
) %>%
  dplyr::mutate(
    class = if_else(
      class == "convalescent",
      paste0(
        "Conv. (",
        sprintf("%.3f", round(summary(cpca_group_sil)$clus.avg.widths[1], 3)),
        ")"),
      if_else(
        class == "DF",
        paste0(
        "DF (",
        sprintf("%.3f", round(summary(cpca_group_sil)$clus.avg.widths[2], 3)),
        ")"),
        paste0("DHF (",
        sprintf("%.3f", round(summary(cpca_group_sil)$clus.avg.widths[3], 3)),
        ")")
      )
    )
  )

cpca_p <- cpca_df %>%
  ggplot(aes(x = cPC1, y = cPC2, colour = class)) +
  geom_point() +
  ggtitle("cPCA") +
  scale_colour_viridis_d(name = "Class",
                         begin = 0.1, end = 0.9) +
  theme_minimal()
cpca_p

# repeat cPCA for n_centers = 2,3,4,5
diff_centers_cpca <- lapply(2:5, function(x) {
  
  # fit cPCA
  scPCA(target, background, center = TRUE, penalties = 0, n_centers = x,
        max_iter = 1000)
  })
```

### cPCA Number of Centers

```{r cpca_diff_centers, fig.retina=3, fig.width=8, fig.height=8}
# this is done one-by-one to ensure the plots' axes are comparable

# two centers:
# compute the average silhouette widths
cpca_2_group_sil <- silhouette(group_mem, dist(diff_centers_cpca[[1]]$x))
cpca_2_df <- diff_centers_cpca[[1]]$x %>%
  as.data.frame() %>%
  dplyr::mutate(class = dengue_class) %>%
  dplyr::mutate(
    class = if_else(
      class == "convalescent",
      paste0(
        "Conv. (",
        sprintf("%.3f", round(summary(cpca_2_group_sil)$clus.avg.widths[1], 3)),
        ")"),
      if_else(
        class == "DF",
        paste0(
        "DF (",
        sprintf("%.3f", round(summary(cpca_2_group_sil)$clus.avg.widths[2], 3)),
        ")"),
        paste0("DHF (",
        sprintf("%.3f", round(summary(cpca_2_group_sil)$clus.avg.widths[3], 3)),
        ")")
      )
    )
  )

colnames(cpca_2_df) <- c("cPC1", "cPC2", "label")

# plot the results
p_cpca_2 <- ggplot(cpca_2_df, aes(x = cPC1, y = -cPC2, colour = label)) +
  geom_point() +
  ggtitle("Two Clusters") +
  xlab("cPC1") +
  ylab("cPC2") +
  scale_colour_viridis_d(name = "Class",
                         begin = 0.1, end = 0.9) +
  theme_minimal()

# three centers:
cpca_3_group_sil <- silhouette(group_mem, dist(diff_centers_cpca[[2]]$x))
cpca_3_df <- diff_centers_cpca[[2]]$x %>%
  as.data.frame() %>%
  dplyr::mutate(class = dengue_class) %>%
  dplyr::mutate(
    class = if_else(
      class == "convalescent",
      paste0(
        "Conv. (",
        sprintf("%.3f", round(summary(cpca_3_group_sil)$clus.avg.widths[1], 3)),
        ")"),
      if_else(
        class == "DF",
        paste0(
        "DF (",
        sprintf("%.3f", round(summary(cpca_3_group_sil)$clus.avg.widths[2], 3)),
        ")"),
        paste0("DHF (",
        sprintf("%.3f", round(summary(cpca_3_group_sil)$clus.avg.widths[3], 3)),
        ")")
      )
    )
  )

colnames(cpca_3_df) <- c("cPC1", "cPC2", "label")

# plot the results
p_cpca_3 <- ggplot(cpca_3_df, aes(x = -cPC1, y = cPC2, colour = label)) +
  geom_point() +
  ggtitle("Three Clusters") +
  xlab("cPC1") +
  ylab("cPC2") +
  scale_colour_viridis_d(name = "Class",
                         begin = 0.1, end = 0.9) +
  theme_minimal()

# four centers:
cpca_4_group_sil <- silhouette(group_mem, dist(diff_centers_cpca[[3]]$x))
cpca_4_df <- diff_centers_cpca[[3]]$x %>%
  as.data.frame() %>%
  dplyr::mutate(class = dengue_class)  %>%
  dplyr::mutate(
    class = if_else(
      class == "convalescent",
      paste0(
        "Conv. (",
        sprintf("%.3f", round(summary(cpca_4_group_sil)$clus.avg.widths[1], 3)),
        ")"),
      if_else(
        class == "DF",
        paste0(
        "DF (",
        sprintf("%.3f", round(summary(cpca_4_group_sil)$clus.avg.widths[2], 3)),
        ")"),
        paste0("DHF (",
        sprintf("%.3f", round(summary(cpca_4_group_sil)$clus.avg.widths[3], 3)),
        ")")
      )
    )
  )

colnames(cpca_4_df) <- c("cPC1", "cPC2", "label")

# plot the results
p_cpca_4 <- ggplot(cpca_4_df, aes(x = -cPC1, y = cPC2, colour = label)) +
  geom_point() +
  ggtitle("Four Clusters") +
  xlab("cPC1") +
  ylab("cPC2") +
  scale_colour_viridis_d(name = "Class",
                         begin = 0.1, end = 0.9) +
  theme_minimal()

# five centers:
cpca_5_group_sil <- silhouette(group_mem, dist(diff_centers_cpca[[4]]$x))
cpca_5_df <- diff_centers_cpca[[4]]$x %>%
  as.data.frame() %>%
  dplyr::mutate(class = dengue_class)  %>%
  dplyr::mutate(
    class = if_else(
      class == "convalescent",
      paste0(
        "Conv. (",
        sprintf("%.3f", round(summary(cpca_5_group_sil)$clus.avg.widths[1], 3)),
        ")"),
      if_else(
        class == "DF",
        paste0(
        "DF (",
        sprintf("%.3f", round(summary(cpca_5_group_sil)$clus.avg.widths[2], 3)),
        ")"),
        paste0("DHF (",
        sprintf("%.3f", round(summary(cpca_5_group_sil)$clus.avg.widths[3], 3)),
        ")")
      )
    )
  )

colnames(cpca_5_df) <- c("cPC1", "cPC2", "label")

# plot the results
p_cpca_5 <- ggplot(cpca_5_df, aes(x = cPC1, y = -cPC2, colour = label)) +
  geom_point() +
  ggtitle("Five Clusters") +
  xlab("cPC1") +
  ylab("cPC2") +
  scale_colour_viridis_d(name = "Class",
                         begin = 0.1, end = 0.9) +
  theme_minimal()

# combine the 4 plots
ggarrange(p_cpca_2, p_cpca_3, p_cpca_4, p_cpca_5,
          ncol = 2, nrow = 2)
```


## scPCA

```{r scpca}
# perform scpca (takes a while to run, so load from saved file)
# see cluster_files folder
load(file = here("analyses/dengue_data/cluster_files/scpca.Rdata"))

# compute the average silhouette widths
scpca_group_sil <- silhouette(group_mem, dist(dengue_scpca$x))

# plot the 2D representation
scpca_df <- data.frame(
  scPC1 = dengue_scpca$x[, 1],
  scPC2 = dengue_scpca$x[, 2],
  class = dengue_class
) %>%
  dplyr::mutate(
    class = if_else(
      class == "convalescent",
      paste0(
        "Conv. (",
        sprintf("%.3f", round(summary(scpca_group_sil)$clus.avg.widths[1], 3)),
        ")"),
      if_else(
        class == "DF",
        paste0(
        "DF (",
        sprintf("%.3f", round(summary(scpca_group_sil)$clus.avg.widths[2], 3)),
        ")"),
        paste0("DHF (",
        sprintf("%.3f", round(summary(scpca_group_sil)$clus.avg.widths[3], 3)),
        ")")
      )
    )
  )

scpca_p <- scpca_df %>%
  ggplot(aes(x = scPC1, y = scPC2, colour = class)) +
  geom_point() +
  ggtitle("scPCA") +
  scale_colour_viridis_d(name = "Class",
                         begin = 0.1, end = 0.9) +
  theme_minimal()
scpca_p
```

```{r save-dengue-cpca-centers}
ggsave(
  filename = here("manuscript", "figures", "dengue_cpca_centers.png"),
  units = "mm",
  width = 178,
  height = 100,
  scale = 1.5,
  dpi = 350
)
```


### scPCA Number of Centers

```{r scpca_diff_centers, fig.retina=3, fig.width=8, fig.height=8}
# this is done one-by-one to ensure the plots' axes are comparable
load(file = here("analyses/dengue_data/cluster_files/diff_centers_scpca.RData"))

# two centers:
scpca_2_group_sil <- silhouette(group_mem, dist(diff_centers_scpca[[1]]$x))
scpca_2_df <- diff_centers_scpca[[1]]$x %>%
  as.data.frame() %>%
  dplyr::mutate(class = dengue_class) %>%
  dplyr::mutate(
    class = if_else(
      class == "convalescent",
      paste0(
        "Conv. (",
        sprintf("%.3f", round(summary(scpca_2_group_sil)$clus.avg.widths[1], 3)),
        ")"),
      if_else(
        class == "DF",
        paste0(
        "DF (",
        sprintf("%.3f", round(summary(scpca_2_group_sil)$clus.avg.widths[2], 3)),
        ")"),
        paste0("DHF (",
        sprintf("%.3f", round(summary(scpca_2_group_sil)$clus.avg.widths[3], 3)),
        ")")
      )
    )
  )

colnames(scpca_2_df) <- c("scPC1", "scPC2", "class")

# plot the results
p_scpca_2 <- ggplot(scpca_2_df, aes(x = scPC1, y = scPC2, colour = class)) +
  geom_point() +
  ggtitle("Two Clusters") +
  xlab("scPC1") +
  ylab("scPC2") +
  scale_colour_viridis_d(name = "Class",
                         begin = 0.1, end = 0.9) +
  theme_minimal()

# three centers:
scpca_3_group_sil <- silhouette(group_mem, dist(diff_centers_scpca[[2]]$x))
scpca_3_df <- diff_centers_scpca[[2]]$x %>%
  as.data.frame() %>%
  dplyr::mutate(class = dengue_class) %>%
  dplyr::mutate(
    class = if_else(
      class == "convalescent",
      paste0(
        "Conv. (",
        sprintf("%.3f", round(summary(scpca_3_group_sil)$clus.avg.widths[1], 3)),
        ")"),
      if_else(
        class == "DF",
        paste0(
        "DF (",
        sprintf("%.3f", round(summary(scpca_3_group_sil)$clus.avg.widths[2], 3)),
        ")"),
        paste0("DHF (",
        sprintf("%.3f", round(summary(scpca_3_group_sil)$clus.avg.widths[3], 3)),
        ")")
      )
    )
  )

colnames(scpca_3_df) <- c("scPC1", "scPC2", "class")

# plot the results
p_scpca_3 <- ggplot(scpca_3_df, aes(x = scPC1, y = scPC2, colour = class)) +
  geom_point() +
  ggtitle("Three Clusters") +
  xlab("scPC1") +
  ylab("scPC2") +
  scale_colour_viridis_d(name = "Class",
                         begin = 0.1, end = 0.9) +
  theme_minimal()

# four centers:
scpca_4_group_sil <- silhouette(group_mem, dist(diff_centers_scpca[[3]]$x))
scpca_4_df <- diff_centers_scpca[[3]]$x %>%
  as.data.frame() %>%
  dplyr::mutate(class = dengue_class) %>%
  dplyr::mutate(
    class = if_else(
      class == "convalescent",
      paste0(
        "Conv. (",
        sprintf("%.3f", round(summary(scpca_4_group_sil)$clus.avg.widths[1], 3)),
        ")"),
      if_else(
        class == "DF",
        paste0(
        "DF (",
        sprintf("%.3f", round(summary(scpca_4_group_sil)$clus.avg.widths[2], 3)),
        ")"),
        paste0("DHF (",
        sprintf("%.3f", round(summary(scpca_4_group_sil)$clus.avg.widths[3], 3)),
        ")")
      )
    )
  )

colnames(scpca_4_df) <- c("scPC1", "scPC2", "class")

# plot the results
p_scpca_4 <- ggplot(scpca_4_df, aes(x = scPC1, y = scPC2, colour = class)) +
  geom_point() +
  ggtitle("Four Clusters") +
  xlab("scPC1") +
  ylab("scPC2") +
  scale_colour_viridis_d(name = "Class",
                         begin = 0.1, end = 0.9) +
  theme_minimal()

# five centers:
scpca_5_group_sil <- silhouette(group_mem, dist(diff_centers_scpca[[4]]$x))
scpca_5_df <- diff_centers_scpca[[4]]$x %>%
  as.data.frame() %>%
  dplyr::mutate(class = dengue_class) %>%
  dplyr::mutate(
    class = if_else(
      class == "convalescent",
      paste0(
        "Conv. (",
        sprintf("%.3f", round(summary(scpca_5_group_sil)$clus.avg.widths[1], 3)),
        ")"),
      if_else(
        class == "DF",
        paste0(
        "DF (",
        sprintf("%.3f", round(summary(scpca_5_group_sil)$clus.avg.widths[2], 3)),
        ")"),
        paste0("DHF (",
        sprintf("%.3f", round(summary(scpca_5_group_sil)$clus.avg.widths[3], 3)),
        ")")
      )
    )
  )

colnames(scpca_5_df) <- c("sscPC1", "scPC2", "class")

# plot the results
p_scpca_5 <- ggplot(scpca_5_df, aes(x = sscPC1, y = -scPC2, colour = class)) +
  geom_point() +
  ggtitle("Five Clusters") +
  xlab("scPC1") +
  ylab("scPC2") +
  scale_colour_viridis_d(name = "Class",
                         begin = 0.1, end = 0.9) +
  theme_minimal()

# combine the 4 plots
ggarrange(p_scpca_2, p_scpca_3, p_scpca_4, p_scpca_5,
          ncol = 2, nrow = 2)
```


```{r svae-dengue-scpca-centers}
ggsave(
  filename = here("manuscript", "figures", "dengue_scpca_centers.png"),
  units = "mm",
  width = 178,
  height = 100,
  scale = 1.5,
  dpi = 350
)
```


### Genes With Non-Zero Loadings

Tables included in the appendix of the manuscript.

```{r map_genes}
# match the affymetrix code to the gene symbol
gene_df <- target %>% colnames %>% tibble()
colnames(gene_df) <- "affy_code"
gene_df <- gene_df %>%
  left_join(gene_names, by = c("affy_code" = "ID")) %>%
  select(affy_code, `Gene Symbol`, `Gene Title`)

# extract the gene symbols that have non-zero loadings in the scPC
rotation_df <- dengue_scpca$rotation %>%
  as_tibble %>%
  dplyr::mutate(
    affy_code = rownames(dengue_scpca$rotation)
  ) %>%
  dplyr::left_join(gene_df, by = "affy_code")
colnames(rotation_df) <- c("scPCA1", "scPCA2", "affy_code", "gene_sym",
                           "gene_title")

# print the table for the supp info
gene_scpca1 <- rotation_df %>%
  dplyr::filter(scPCA1 != 0) %>%
  select(gene_sym, gene_title, scPCA1) %>%
  xtable(
    caption = "Genes with non-zero loadings in first scPCA loadings vector.",
    digits = 4
  )

gene_scpca2 <- rotation_df %>%
  dplyr::filter(scPCA2 != 0) %>%
  select(gene_sym, gene_title, scPCA2) %>%
  xtable(
    caption = "Genes with non-zero loadings in second scPCA loadings vector.",
    digits = 4
  )
```

### Comparison of cPCA and scPCA Loadings

```{r comp_loadings, warning=FALSE, message=FALSE}
# compute the relative absolute loading ratios
cpca_loads_df <- dengue_cpca$rotation %>%
  as.data.frame() %>%
  dplyr::mutate(
    V1 = abs(V1),
    V2 = abs(V2),
    V1 = (V1-min(V1))/(max(V1)-min(V1)),
    V2 = (V2-min(V2))/(max(V2)-min(V2))
  )
scpca_loads_df <- dengue_scpca$rotation %>%
  as.data.frame() %>%
  dplyr::mutate(
    V1 = abs(V1),
    V2 = abs(V2),
    V1 = (V1-min(V1))/(max(V1)-min(V1)),
    V2 = (V2-min(V2))/(max(V2)-min(V2))
  )

# create the loadings comparison plot
load_diff_df <- bind_rows(
  cpca_loads_df,
  scpca_loads_df
) %>%
  dplyr::mutate(
    sparse = c(rep("0", ncol(target)), rep("1", ncol(target))),
    gene = rep(1:ncol(target), 2)
  )

# order the genes based on decreasing rel abs cpca loadings
ord_1_genes <- factor(load_diff_df$gene,
                      levels = sort(load_diff_df$V1[1:500], decreasing = TRUE,
                                    index.return = TRUE)$ix)
ord_2_genes <- factor(load_diff_df$gene,
                      levels = sort(load_diff_df$V2[1:500], decreasing = TRUE,
                                    index.return = TRUE)$ix)
load_diff_df <- load_diff_df %>%
  dplyr::mutate(
    gene1 = ord_1_genes,
    gene2 = ord_2_genes
  )

colnames(load_diff_df) <- c("comp1", "comp2", "sparse", "gene",
                            "gene1", "gene2")

p1 <- load_diff_df %>%
  ggplot(aes(y = comp1, x = gene1, fill = sparse, alpha = sparse)) +
    geom_bar(stat = "identity", position = position_identity(), width = 1) +
    xlab("Gene") +
    ylab("Standardized Abs.\n Loadings") +
    ggtitle("First Loadings Vector Comparison") +
    scale_fill_viridis_d(name = "Method",
                         labels = c("cPCA", "scPCA"),
                         begin = 0.9, end = 0.1) +
    scale_alpha_discrete(range = c(0.5, 1),
                         guide = FALSE) +
    theme_minimal() +
    theme(panel.grid.major = element_blank(),
          panel.grid.minor = element_blank(),
          axis.text.x=element_blank(),
          axis.ticks.x=element_blank())

p2 <- load_diff_df %>%
  ggplot(aes(y = comp2, x = gene2, fill = sparse, alpha = sparse)) +
    geom_bar(stat = "identity", position = position_identity(), width = 1) +
    xlab("Genes") +
    ylab("Standardized Abs.\n Loadings") +
    ggtitle("Second Loadings Vector Comparison") +
    scale_fill_viridis_d(name = "Method",
                         labels = c("cPCA", "scPCA"),
                         begin = 0.9, end = 0.1) +
    scale_alpha_discrete(range = c(0.5, 1),
                         guide = FALSE) +
    theme_minimal() +
    theme(panel.grid.major = element_blank(),
          panel.grid.minor = element_blank(),
          axis.text.x=element_blank(),
          axis.ticks.x=element_blank())


loadings_comp <- annotate_figure(
  ggarrange(p1, p2, nrow = 1, ncol = 2,
            common.legend = TRUE, legend = "right")
)
loadings_comp
```

## cPCA (Tuned with Cross-Validation)

3-fold cross-validation was used to tune the contrastive parameter.

```{r cPCA_cv}
# see cluster_files folder
load(file = here("analyses/dengue_data/cluster_files/cpca_cv.Rdata"))

# compute the average silhouette widths
cpca_cv_group_sil <- silhouette(group_mem, dist(dengue_cpca_cv$x))

# plot the 2D representation
cpca_cv_df <- data.frame(
  cPC1 = -dengue_cpca_cv$x[, 1],
  cPC2 = dengue_cpca_cv$x[, 2],
  class = dengue_class
) %>%
  dplyr::mutate(
    class = if_else(
      class == "convalescent",
      paste0(
        "Conv. (",
        sprintf("%.3f", round(summary(cpca_cv_group_sil)$clus.avg.widths[1], 3)),
        ")"),
      if_else(
        class == "DF",
        paste0(
        "DF (",
        sprintf("%.3f", round(summary(cpca_cv_group_sil)$clus.avg.widths[2], 3)),
        ")"),
        paste0("DHF (",
        sprintf("%.3f", round(summary(cpca_cv_group_sil)$clus.avg.widths[3], 3)),
        ")")
      )
    )
  )

cpca_cv_p <- cpca_cv_df %>%
  ggplot(aes(x = cPC1, y = cPC2, colour = class)) +
  geom_point() +
  ggtitle("cPCA") +
  scale_colour_viridis_d(name = "Class",
                         begin = 0.1, end = 0.9) +
  theme_minimal()
cpca_cv_p
```



## scPCA (Tuned with Cross-Validation)

3-fold cross-validation was used to tune the hyperparameters.

```{r scPCA_cv}
# see cluster_files folder
load(file = here("analyses/dengue_data/cluster_files/scpca_cv.Rdata"))

# compute the average silhouette widths
scpca_cv_group_sil <- silhouette(group_mem, dist(dengue_scpca_cv$x))

# plot the 2D representation
scpca_cv_df <- data.frame(
  scPC1 = dengue_scpca_cv$x[, 1],
  scPC2 = dengue_scpca_cv$x[, 2],
  class = dengue_class
) %>%
  dplyr::mutate(
    class = if_else(
      class == "convalescent",
      paste0(
        "Conv. (",
        sprintf("%.3f", round(summary(scpca_cv_group_sil)$clus.avg.widths[1], 3)),
        ")"),
      if_else(
        class == "DF",
        paste0(
        "DF (",
        sprintf("%.3f", round(summary(scpca_cv_group_sil)$clus.avg.widths[2], 3)),
        ")"),
        paste0("DHF (",
        sprintf("%.3f", round(summary(scpca_cv_group_sil)$clus.avg.widths[3], 3)),
        ")")
      )
    )
  )

scpca_cv_p <- scpca_cv_df %>%
  ggplot(aes(x = scPC1, y = scPC2, colour = class)) +
  geom_point() +
  ggtitle("scPCA") +
  scale_colour_viridis_d(name = "Class",
                         begin = 0.1, end = 0.9) +
  theme_minimal()
scpca_cv_p
```


```{r save-dengue-data-cv-plot}
comb_cv_p <- ggarrange(
  cpca_cv_p,
  scpca_cv_p,
  nrow = 1,
  ncol = 2
)
ggsave(
  filename = here("manuscript", "figures", "dengue_cv_results.png"),
  units = "mm",
  width = 178,
  height = 60,
  dpi = 320,
  scale = 1.5,
  plot = comb_cv_p
)
```

## t-SNE

### With Initial PCA Step

```{r tsne_with_pca}
# get tsne representation
dengue_tsne_pca <- Rtsne(target, perplexity = 8, pca = TRUE, max_iter = 1000,
                         theta = 0)

# t-SNE silhouette width
tsne_pca_group_sil <- silhouette(group_mem, dist(dengue_tsne_pca$Y))

# plot the first two axes
tsne_pca_df <- data.frame(
  TSNE1 = dengue_tsne_pca$Y[, 1],
  TSNE2 = dengue_tsne_pca$Y[, 2],
  class = dengue_class
) %>%
  dplyr::mutate(
    class = if_else(
      class == "convalescent",
      paste0(
        "Conv. (",
        sprintf("%.3f", round(summary(tsne_pca_group_sil)$clus.avg.widths[1], 3)),
        ")"),
      if_else(
        class == "DF",
        paste0(
        "DF (",
        sprintf("%.3f", round(summary(tsne_pca_group_sil)$clus.avg.widths[2], 3)),
        ")"),
        paste0("DHF (",
        sprintf("%.3f", round(summary(tsne_pca_group_sil)$clus.avg.widths[3], 3)),
        ")")
      )
    )
  )


tsne_pca_p <- tsne_pca_df %>%
  ggplot(aes(x = TSNE1, y = TSNE2, colour = class)) +
  geom_point() +
  xlab("tSNE1") +
  ylab("tSNE2") +
  ggtitle("t-SNE") +
  scale_colour_viridis_d(name = "Class",
                         begin = 0.1, end = 0.9) +
  theme_minimal()
tsne_pca_p
```

### Without Initial PCA Step

```{r tsne}
# get tsne representation
dengue_tsne <- Rtsne(target, perplexity = 8, pca = FALSE, theta = 0)

# t-SNE silhouette width
tsne_group_sil <- silhouette(group_mem, dist(dengue_tsne$Y))

# plot the first two axes
tsne_df <- data.frame(
  TSNE1 = dengue_tsne$Y[, 1],
  TSNE2 = dengue_tsne$Y[, 2],
  class = dengue_class
) %>%
  dplyr::mutate(
    class = if_else(
      class == "convalescent",
      paste0(
        "Conv. (",
        sprintf("%.3f", round(summary(tsne_group_sil)$clus.avg.widths[1], 3)),
        ")"),
      if_else(
        class == "DF",
        paste0(
        "DF (",
        sprintf("%.3f", round(summary(tsne_group_sil)$clus.avg.widths[2], 3)),
        ")"),
        paste0("DHF (",
        sprintf("%.3f", round(summary(tsne_group_sil)$clus.avg.widths[3], 3)),
        ")")
      )
    )
  )


tsne_p <- tsne_df %>%
  ggplot(aes(x = TSNE1, y = TSNE2, colour = class)) +
  geom_point() +
  xlab("tSNE1") +
  ylab("tSNE2") +
  ggtitle("t-SNE") +
  scale_colour_viridis_d(name = "Class",
                         begin = 0.1, end = 0.9) +
  theme_minimal()
tsne_p
```

```{r dengue_t-sne_plot}
ggsave(
  filename = here("manuscript", "figures", "dengue_tsne.png"),
  units = "mm",
  width = 100,
  height = 70,
  scale = 1.5,
  dpi = 350
)
```


Since both initializations produced qualitatively identical embeddings, t-SNE
without the initial PCA step is considered in the manuscript.

## UMAP

```{r umap}
# set the parameters
params <- umap.defaults
params$n_neighbors <- 15
params$min_dist <- 0.2

# get the umap representation
dengue_umap <- umap(target, config = params)

# umap silhouette width
umap_group_sil <- silhouette(group_mem, dist(dengue_umap$layout))

# plot the first two axes
umap_df <- dengue_umap$layout %>%
  as.data.frame() %>%
  dplyr::mutate(class = dengue_class) %>%
  dplyr::mutate(
    class = if_else(
      class == "convalescent",
      paste0(
        "Conv. (",
        sprintf("%.3f", round(summary(umap_group_sil)$clus.avg.widths[1], 3)),
        ")"),
      if_else(
        class == "DF",
        paste0(
        "DF (",
        sprintf("%.3f", round(summary(umap_group_sil)$clus.avg.widths[2], 3)),
        ")"),
        paste0("DHF (",
        sprintf("%.3f", round(summary(umap_group_sil)$clus.avg.widths[3], 3)),
        ")")
      )
    )
  )
  
colnames(umap_df) <- c("UMAP1", "UMAP2", "class")

umap_p <- umap_df %>%
  ggplot(aes(x = UMAP1, y = UMAP2, colour = class)) +
  geom_point() +
  ggtitle("UMAP") +
  scale_colour_viridis_d(name = "Class",
                         begin = 0.1, end = 0.9) +
  theme_minimal()
umap_p
```


# Combined Plots

```{r comb_plots, fig.retina=3, fig.width=8, fig.height=8}
comb_plot <- ggarrange(
  ggarrange(pca_p, umap_p, cpca_p, scpca_p,
            ncol = 2, nrow = 2),
  loadings_comp,
  nrow = 2,
  labels = "AUTO",
  heights = c(2, 1)
)
comb_plot
```

```{r save_dengue_data_plot, eval=FALSE}
ggsave(
  filename = here("manuscript", "figures", "dengue_results.png"),
  width = 178,
  height = 140,
  units = "mm",
  dpi = 350,
  plot = comb_plot,
  scale = 1.5
)
```


